\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Local Adaptive Optimization of Time Step}

### 2013-11-05 report/article.tex
Final report done

\usepackage{titlesec} % Allows customization of titles
\renewcommand\thesection{\Roman{section}} % Roman numerals for the sections
\renewcommand\thesubsection{\Roman{subsection}} % Roman numerals for subsections
\titleformat{\section}[block]{\large\scshape\centering}{\thesection.}{1em}{} % Change the look of the section titles
\titleformat{\subsection}[block]{\large}{\thesubsection.}{1em}{} % Change the look of the section titles

@@

\author{
\large
\textsc{Malte Stær Nissen}\thanks{Thanks to my supervisor Sune Darkner for
helping me realize this project}\\[2mm] % Your name
\normalsize University of Copenhagen \\ % Your institution
\normalsize \href{mailto:nissen@diku.dk}{nissen@diku.dk} % Your email address
\vspace{-5mm}
}
\date{}

@@

We will in this report
study possible approaches for application of LAOT. In order to keep the focus
on possible LAOT methods and remove unneccessary complexities, we will study
the simple 1 dimensional simulation of a mass-spring system and try to apply
LAOT for it. Afterwards we will shortly discuss the application of our gained
knowledge to the domain of hyperelastic material simulation and simulations in
general.

@@

Recently newer and faster methods with local timestepping have been developed
on the basis of nonlinear PDEs and the work performed in this field in
the 80s and 90s. First the multi resolution (MR) schemes for creating
space-adaptive discretizations and refinements of these were developed, see
\cite{Berger:1984}. Then multiple MR based schemes were developed. Domingues
et al. \cite{Domingues:2008} is one of the more recent of these MR schemes.
They describe a local scale-dependent timestepping for a space-adaptive multi
resolution scheme using the finite volume method in order to obtain speed-up
using larger time-steps without violating a defined stability constraints,
which is essentially the same motivation as this report. The method is based
on an explicit Runge-Kutta method of second order. As expected the timestep
size is imposed by a stability condition of the explicit Runge-Kutta on the
finest scale, which increases with the scale of the mesh and hence we are able
to increase the timestep as well without violating the stability condition.

@@

\section{Methods}
\label{sec:methods}

With no prerequisites within the field of mass simulations we will first
consider the 1D problem of simulating a mass-spring system using the explicit
(forward) Euler method.

Second we will propose and test out three different methods for applying LAOT
to the simulation.

Third and last we will discussion the results and how we could apply our
knowledge of LAOT for hyperelastic material.

\subsection{1D mass-spring system} Mass-spring systems are systems simulating
a set of vertices interconnected by springs in various ways to obtain
desirable physical properties. The motion of the springs and vertices are
then simulated as a function of the time elapsed. Examples of application of
mass-spring systems are simulation of cloth or hair. When performing simulations
we want to avoid a switch of position of spring-connected vertices caused by the
timestep size of the simulation, since this invalidates our discretization of
the domain. Furthermore we want to perform as few timesteps as possible in order
to get to the final simulation state by using as big timesteps as possible
without affecting the outcome of the simulation.

When simulating the mass-spring system we perform the following steps:
\begin{enumerate} \item Initialize system \item Calculate forces \item
Calculate derrivatives \item Update positions and velocities \item Update time
\item Go to step 2 if the simulation has not reached it's maximum simulation
time. \end{enumerate} We now describe the details of each of the steps of
the simulation.

Our 1D mass-spring system consists of a set $V$ of vertices and a set
$S$ of springs. Each vertex $v \in V$ has a specified position $p_v$,
mass $m_v$, force $F_v$ and velocity $u_v$, and each spring $s \in S$
has two endpoints vertices $p_{f,s} \in V$ and $p_{t,s} \in V \setminus
p_{f,s}$, a damping factor $d_s$, a stiffness constant $K_s$ and a rest length
$l_{0,s}$. All these variables and constants combined describe the state
of the mass-spring system at each time $t$ during a simulation. All these
variables are initialized in step 1.

The total force of a spring is the sum of the spring force $F_{h,d}$
calculated by using Hooke's Law (see \cite[p.~439]{Young:2010})
and the damping force $F_{C,d}$ using a simple damping method (see
\cite[p.~457]{Young:2010}): \begin{align} \nonumber l_d &= p_{l,d} -
p_{r,d} \\ \label{eq:hooke} F_{h,d} &= -k_d x_d = -k_d (l_d - l_{0,d}) \\
\label{eq:damping} F_{C,d} &= -C_d \frac{\partial x_d}{ \partial t} = - C_d
\frac{\left(v_{l,d} - v_{r,d}\right) l_d}{\abs{l_d}} \\ F_d &= F_{h,d} +
F_{C,d} \end{align} \todo{Maybe add direction correction: $F_d \cdot (-l_d
/ \abs{l_d})$} in which $p_{l,d}$ and $p_{r,d}$ are the positions of the
left and right endpoints of the spring $d$ respectively, and $v_{l,d}$ and
$v_{r,d}$ are the velocities of the left and right endpoint vertices of the
spring $d$ respectively, and $F_d$ is the total force of spring $d$.

When calculating the forces applied by each spring, we add the total spring
force $F_d$ to the left endpoint of the spring and subtract $F_d$ from the
right endpoint since both endpoints are affected equally in opposite direction
by the spring $d$ connecting them. These spring forces are calculated in step
2 of the simulation.

We now know how to calculate the force of the spring system and apply it to
the vertices. In order to calculate the derrivatives of the system (step 3) we
use Newton's second law of motion (see \cite[p.~112-115]{Young:2010}) to get
the difference in speed $\Delta u_v$ and position $\Delta p_v$ of each vertex
$v$ using a given timestep size $\Delta t$: \begin{align} \label{eq:delta_u}
\Delta u_v &= \frac{\partial u_v}{\partial t} \Delta t = \frac{F_v}{m_v}
\Delta t \\ \label{eq:delta_p} \Delta p_v &= u \Delta t \end{align}

As mentioned in the beginning of this section we have chosen to use the
explicit Euler method defined below in equation \ref{eq:explicit_euler}
to calculate the time ``integration'' \cite{Flaherty} when stepping through
the simulation. Given a variable $x$ as a function of the time $t$, the
variable $x$ at the time $t + \Delta t$ is defined as: \begin{align}
\label{eq:explicit_euler} x(t+\Delta t) &= x(t) + \Delta t \frac{\partial
x(t)}{\partial t} \end{align} Applying this method for the timestepping
of our simulation we get the two following position and velocity update
equations using equations \ref{eq:delta_u} and \ref{eq:delta_p}: \begin{align}
p_v(t+\Delta t) &= p_v(t) + \Delta p_v \\ u_v(t+\Delta t) &= u_v(t) + \Delta
u_v \end{align} We use these equations to perform step 4 of the simulation.

Step 5 is a simple update of the time: $t = t + \Delta t$ and step 6 describes
itself.

We have now fully described the simple case of simulating a mass-spring system
using the explicit Euler method and the next subsections describe the three
different methods/modifications for applying LAOT to the simulation just
described. The three methods are using different ideas to implement LAOT. They
all include using some sort of heuristics in order to perform the adaption of
the methods. We developed and tested these heuristics as we developed the
methods (chronologically) and hence most of the heuristics could potentially be
used on all three methods and not only the methods on which they were used. In
case the methods are working succesfully, they could furthermore be combined to
generate ie. fully atomic timestepping (both time and spatial adaptive).

\subsection{Adapt-on-switch}
\label{sec:method_switch}
The first LAOT method we propose is time-adaptive. We perform a simulation
with an initial timestep size. In each step we try to predict whether a timestep
of the same size using the current velocities will result in a switch of the
left and right vertices of any of the springs. If this is the case, we should
try to avoid the switch by decreasing the timestep of the current iteration,
re-calculate the velocity and position changes and resume stepping of the
simulation using the new timestep size. When switches aren't detected anymore
we gradually increase the timestep of the simulation again. We call this
method the adapt-on-switch method.

The switch of two endspoints connected by a spring $d$ can easily be
calculated by checking the length ($p_{l,d} - p_{r,d}$) of the spring before
and after a position update. If the sign of the length differs, the two
points must have switched sides and hence adaption of the timestep should be
performed. The adaption of the timestep can either be done by division of the
current timestep by a constant or by having a finite set of timestep sizes to
choose from.

@@

The heuristics can be chosen to fit the domain of the simulation, but we have
chosen to use the number of steps since last update, $n_{last}$, and the ratio
between the current spring length and the spring length at the last update
$r_{last}$. This should force a bound on the maximal number of timesteps
before update as well as force frequent updates when the spring is affected by a
high acceleration in opposite direction of the current velocity/direction of
change. \todo{Explain length ratio thought}. Further details are explained in
section \ref{sec:experiments_atomic}.

\subsection{Global Estimated Acceleration}
The third and last LAOT method we propose is time-adaptive. The simulations are
done using an initial timestep size and define the timestep of each iteration
based on a heuristic using the acceleration of the springs, and hence we call
the method Global Estimated Acceleration (GEA). We know that the force and hence
acceleration of a spring is somewhat linear close to the spring's rest length
and it ``bends'' off as the spring is streched or contracted further away from
the rest length. Based on this knowledge we will try to make a heuristic
telling us about the needed timestep size at any given point during the
simulation. The smallest needed timestep defines the global timestep for that
given step.

For this simple 1D example we use the inverse of the acceleration
for adapting the timestep and hence the heuristic is defined as:
\begin{align}
    \frac{1}{a} &= \frac{1}{\frac{-kx}{m}} = \frac{m}{-kx} \\
    \label{eq:inverse}
    \Delta t &= \text{min}\left( \frac{1}{\abs{a} + 0.1} \right )
\end{align}
We have added a small value (0.1) in order to avoid the timestep to increase
exponentially towards infinity as $\abs{a}$ approach 0. Furthermore a scaling
of the timestep could be needed depending on the scale of the simulation mesh.
Further details are explained in section \ref{sec:experiments}.

Our though is that for more
complex simulations we are able to make estimates of this type less costly than
performing the actual simulations of a small fixed size timestep.

@@

\section{Experiments, results and discussion}
\label{sec:experiments}

We first give an example of the actual application of LAOT using the
standard simulation of mass spring particles. Figure \ref{fig:std} shows two
simulations of the same set of particles distributed with two different scales
simulated for 250 seconds each. All the springs $s \in S$ are initiated in a
stretched state/position with a length 1.25 times longer than the rest length,
damping and stiffness constants $C_s = 0.5, K_s = 0.5$, and all vertices $v
\in V$ have a mass of $m_v = 30$. Given this initial state all springs should
start constracting since they are in a stretched position and eventually
somewhere further in time stabilize at the rest position since we have added
the damping force to the system.

\ref{fig:std_multi_30fps} shows a simulation of the system described above
using 30 fps and Figure \ref{fig:std_multi_1fps} shows the simulation using 1
fps. We clearly see that using a timestep of 30 fps the simulation is stable
and able to handle the compression of all the small springs around $t = 110$
whereas a switch occurs when simulating at 1 fps. This switch causes ripples
of sideeffects for the rest of the simulation and hence the final result is
incorrect. No switches of the longer springs are however occuring and hence
these pose no problem at 1 fps. We will use this simulation for testing all
three LAOT methods proposed in section \ref{sec:methods}.

@@

%\begin{figure}[H]
%    \begin{subfigure}[t]{0.5\textwidth}
%        \includegraphics[width=\textwidth]{../images/standard_uniform_30fps.pdf}
%        \caption{30 fps}
%        \label{fig:std_uni_30fps}
%    \end{subfigure}
%    \begin{subfigure}[t]{0.5\textwidth}
%        \includegraphics[width=\textwidth]{../images/standard_uniform_1fps.pdf}
%        \caption{1 fps}
%        \label{fig:std_uni_1fps}
%    \end{subfigure}
%    \caption{Standard simulations of particles distributed uniformly using
%    different timestep sizes}
%\end{figure}

\subsection{Adapt-on-switch}
As mentioned in section \ref{sec:method_switch} the adapt-on-switch method
can be implemented using either a relative or a absolute adaption of the
timestep. In our experiments we have tested out both methods defining the
relative adaption to scale the timestep by a factor of $\frac{1}{100}$ and
the absolute solution to set the timestep to $\frac{1}{30}$ when switches are
detected since this generated nice simulations for the standard simulation .
Furthermore the up-scaling for the relative scheme is set to a factor of 10
since we wish to avoid running into new switches immediately.

Figure \ref{fig:switch} shows the result of running simulations
described earlier at 1 fps using the adapt-on-switch scheme, where
the red crosses mark timesteps where a switch is predicted in the
following step and hence the schemes are ``activated''. Figure
\ref{fig:switch_multi_1fps_relative} is simulated using the relative scheme
and figure \ref{fig:switch_multi_1fps_absolute} is simulated using the
absolute scheme. As we clearly see in both figures the schemes fail to correct
the simulations in spite of the switch prediction. Once the first switch
occurs the simulation will change behavior and hence the first red cross marks
the point at which the simulation can be rejected from being correct.

After having made these observations we furthermore tested out a third way of
creating the scheme by predicting how long a timestep we could take before the
switch would occur and then divide this steplength by 2 to get a new timestep
length. This generated infeasibly small timesteps forcing us to stop the
simulation too early.

Taking these observations into account we can safely say that the
adapt-on-switch method/scheme is not working for our simulation. This does
however not mean that the method couldn't work for other domains or other
refinements of the method. The problem with the scheme is that the prediction
of switches is done too late in the simulation and hence once we detect
the switches it's too late to do anything about them given our model of
the mass-spring system. In order for the scheme to work properly it should
therefore be modified to be able to perform the detection of switches earlier
in the simulation.

In the real world the spring force would ``bend'' towards $\pm \infty$ once
the spring is stretched or contracted too much and hence the linear model that
we use doesn't suffice when using the adapt-on-switch scheme. Having used a
more realistic choice of model could possibly have had a positive impact of
the use of this scheme since this would cause the spring force of the spring
in question to grow in order to avoid the switch. \todo{Maybe the section
above should be moved to a discussion section.}

@@

\subsection{Atomic Updating using Heuristics}
\label{sec:experiments_atomic}
The AUH method concists of two heuristics for determining whether an element
should be updated or not: steps since last update $l_{last}$, and the ratio
between the spring length in last update and the current length $r_{last}$.

We first test the method using only the first heuristic which will give
us synchronized updates of all vertices every $r_{last}$ step. Using the
same setup as the standard simulation we perform simulations using the AUH
method with the length ratio heuristic ignored by setting the limit of
the ratio to infinity. We set the timestep to $\Delta t = \frac{1}{10}$
(10 fps) and vary $l_{last}$. Figure \ref{fig:atomic_multi_sbound} shows
the resulting simulations where figures \ref{fig:atomic_multi_10sbound},
\ref{fig:atomic_multi_68sbound}, and \ref{fig:atomic_multi_69sbound} uses
an $l_{last}$ of 10, 68, and 69 respectively.

Simulating using 10 fps and $l_{last} = 10$ gives us the same amount
of velocity updates as the standard simulation using 1 fps (which had
switches), but gives us a nice and smooth simulation with no problems.
As can be seen in the figures \ref{fig:atomic_multi_68sbound} and
\ref{fig:atomic_multi_69sbound} we are able to perform the simulation using
$l_{last} = 68$ without having switches occuring whereas using $l_{last} =
69$ results in switches occuring. In other words we are able to perform the
simulation at 10 fps but with 6.8 seconds between acceleration and velocity
calculations. This is a result of combining the explicit Euler method with
the AUH scheme since the scheme simply extrapolates the position at the last
acceleration and velocity update to a position close in time to the next
update position and then uses this position when updating the acceleration and
velocity again. Had we chosen an alternative time integration method this
advantage would most likely have been eliminated and let us perform the larger
timesteps with the standard simulation.

We now turn our attention towards the second heuristic.

\begin{figure}[H]
    \begin{subfigure}[t]{0.5\textwidth}
        \includegraphics[width=\textwidth]{../images/atomic_multiscale_10fps_10sbound.pdf}
        \caption{$l_{last} = 10$, $r_{last}$ ignored}
        \label{fig:atomic_multi_10sbound}
    \end{subfigure}
    \begin{subfigure}[t]{0.5\textwidth}
        \includegraphics[width=\textwidth]{../images/atomic_multiscale_10fps_68sbound.pdf}
        \caption{$l_{last} = 68$, $r_{last}$ ignored}
        \label{fig:atomic_multi_68sbound}
    \end{subfigure}
    \begin{subfigure}[t]{0.5\textwidth}
        \includegraphics[width=\textwidth]{../images/atomic_multiscale_10fps_69sbound.pdf}
        \caption{$l_{last} = 69$, $r_{last}$ ignored}
        \label{fig:atomic_multi_69sbound}
    \end{subfigure}
    \begin{subfigure}[t]{0.5\textwidth}
        \includegraphics[width=\textwidth]{../images/atomic_multiscale_10fps_both.pdf}
        \caption{$l_{last} = 10, r_{last} = 1.01$}
        \label{fig:atomic_multi_both}
    \end{subfigure}
    \caption{Simulations using the AUH method at 10 fps with various number of
    steps since last update.}
    \label{fig:atomic_multi_sbound}
\end{figure}

\begin{figure}
    \begin{subfigure}[t]{0.5\textwidth}
        \includegraphics[width=\textwidth]{../images/atomic_uniform_10fps_0.pdf}
        \caption{}
        \label{fig:atomic_uniform_10fps_0}
    \end{subfigure}
    \begin{subfigure}[t]{0.5\textwidth}
        \includegraphics[width=\textwidth]{../images/atomic_uniform_10fps_1.pdf}
        \caption{}
        \label{fig:atomic_uniform_10fps_1}
    \end{subfigure}
    \caption{}
    \label{fig:atomic_uniform}
\end{figure}

if ratio then too much dampening caused by lack of updates: Example: Simple with
3 springs

>>>>

\usepackage{titlesec} % Allows customization of titles
%\renewcommand\thesection{\Roman{section}} % Roman numerals for the sections
%\renewcommand\thesubsection{\Roman{subsection}} % Roman numerals for subsections
\titleformat{\section}[block]{\large\scshape\centering}{\thesection.}{1em}{} % Change the look of the section titles
\titleformat{\subsection}[block]{\large}{\thesubsection.}{1em}{} % Change the look of the section titles

@@

\author{
\large
\textsc{Malte Stær Nissen}\thanks{Thanks to my supervisor Sune Darkner for
helping and guiding me through this project}\\[2mm] % Your name
\normalsize University of Copenhagen \\ % Your institution
\normalsize \href{mailto:nissen@diku.dk}{nissen@diku.dk} % Your email address
\vspace{-5mm}
}
\date{}

@@

We will in this report
study three possible approaches/methods for application of LAOT and the
heuristics included in these methods. In order to keep the focus
on possible LAOT methods and remove unneccessary complexities, we will study
the simple 1 dimensional simulation of a mass-spring system and try to apply
LAOT for it. Afterwards we will test our methods and discuss on the results and
the applied heuristics.

@@

Recently newer and faster methods with local timestepping have been developed
on the basis of nonlinear PDEs and the work performed in this field in
the 80s and 90s. First the multi resolution (MR) schemes for creating
space-adaptive discretizations and refinements of these were developed, see
\cite{Berger:1984}. Then multiple schemes based on these MR schemes were
developed. Domingues
et al. \cite{Domingues:2008} is one of the more recent of these MR schemes.
They describe a local scale-dependent timestepping for a space-adaptive multi
resolution scheme using the finite volume method in order to obtain speed-up
using larger time-steps without violating defined stability constraints,
which is essentially the same motivation as this report. The method is based
on an explicit Runge-Kutta method of second order. As expected the timestep
size is imposed by a stability condition of the explicit Runge-Kutta on the
finest scale, which increases with the scale of the mesh and hence we are able
to increase the timestep as well without violating the stability condition.

@@

\section{Methods}
\label{sec:methods}
With no prerequisites within the field of mass simulations we will first
consider the 1D problem of simulating a mass-spring system using the explicit
(forward) Euler method.

Second we will propose three different methods and heuristics for applying
LAOT to the simulation. These methods will be tested and the results will be
discussed in section \ref{sec:experiments} and \ref{sec:future_work}.

Both the standard 1D mass-spring system simulator and simulators for each
of the three LAOT methods have been implemented as described in this section.
All figures are results of the implemented simulators.

\subsection{1D mass-spring system}
Mass-spring systems are systems simulating a set of vertices interconnected by
springs in various ways to obtain desirable physical properties. The motion
of the springs and vertices are simulated as a function of the time elapsed.
Examples of application of mass-spring systems are simulation of cloth or
hair. When performing simulations we want to avoid a switch of position of
spring-connected vertices caused by the timestep size of the simulation, since
this invalidates our discretization of the domain. Furthermore we want to reach
the final simulation state having performed as few computations as possible.
This is done by having a good balance between the computational load of each
step and having as large steps as possible without affecting the simulation
outcome.

When simulating the mass-spring system we perform the following six steps:
\begin{enumerate}
    \item Initialize system
    \item Calculate forces
    \item Calculate derrivatives
    \item Update positions and velocities
    \item Update time
    \item Go to step 2 if the simulation has not reached it's maximum
simulation time.
\end{enumerate}
We now go through the details of each of the steps of the simulation.

Our 1D mass-spring system consists of a set $V$ of vertices and a set
$S$ of springs. Each vertex $v \in V$ has a specified position $p_v$,
mass $m_v$, force $F_v$ and velocity $u_v$. Each spring $s \in S$
has two endpoints vertices $p_{f,s} \in V$ and $p_{t,s} \in V \setminus
p_{f,s}$, a damping factor $C_s$, a stiffness constant $K_s$ and a rest length
$l_{0,s}$. The set of all these variables and constants for both vertices and
springs describes the state
of the mass-spring system at each time $t$ during a simulation. The entire set
of variables and constants is initialized in step 1 as well as a variable
$t_{stop}$ defining the duration/stopping time of the simulation.

The total force of a spring $s \in S$ is the sum of the spring force $F_{h,s}$
calculated by using Hooke's Law (see \cite[p.~439]{Young:2010})
and the damping force $F_{C,s}$ using a simple damping method (see
\cite[p.~457]{Young:2010}):
\begin{align}
    \nonumber l_d &= p_{l,s} - p_{r,s} \\
    \label{eq:hooke}
    F_{h,s} &= -k_s x_s = -k_s (l_s - l_{0,s}) \\
    \label{eq:damping}
    F_{C,s} &= -C_s \frac{\partial x_s}{ \partial t} = - C_s
        \frac{\left(v_{l,s} - v_{r,s}\right) l_s}{\abs{l_d}} \\
    F_s &= \left( F_{h,s} + F_{C,s} \right) \frac{-l_s}{\abs{l_s}}
\end{align}
In which $p_{l,s}$ and $p_{r,s}$ are the positions of the
left and right endpoints of the spring $s$ respectively, $v_{l,s}$ and
$v_{r,s}$ are the velocities of the left and right endpoint vertices of the
spring $s$ respectively, and $F_s$ is the total force of spring $s$. Multiplying
the total spring force by $\frac{-l_s}{\abs{l_s}}$ allows us to correct for the
direction of the spring getting the final direction-oriented total spring force
$F_s$.

When calculating the forces applied by each spring, we subtract the total spring
force $F_d$ from the left endpoint of the spring and add $F_d$ to the
right endpoint since both endpoints are affected equally in opposite direction
by the spring $s$ connecting them. These spring forces are calculated in step
2 of the simulation.

We now know how to calculate the spring forces of the spring system and apply them to
the vertices. In order to calculate the first order derrivatives of the system (step 3) we
use Newton's second law of motion (see \cite[p.~112-115]{Young:2010}) to get
the difference in speed $\Delta u_v$ and position $\Delta p_v$ of each vertex
$v$ using a given timestep size $\Delta t$:
\begin{align}
    \label{eq:delta_u}
    \Delta u_v &= \frac{\partial u_v}{\partial t} \Delta t = \frac{F_v}{m_v}
\Delta t \\
    \label{eq:delta_p} \Delta p_v &= u \Delta t
\end{align}

As mentioned in the beginning of this section we have chosen to use the
explicit Euler method (see \cite{Flaherty}) defined below in equation \ref{eq:explicit_euler}
to calculate the time ``integration'' when stepping through
the simulation. Given a variable $x$ as a function of the time $t$, the
variable $x$ at the time $t + \Delta t$ is defined as:
\begin{align}
    \label{eq:explicit_euler}
    x(t+\Delta t) &= x(t) + \Delta t \frac{\partial x(t)}{\partial t}
\end{align}
Applying this method for the timestepping of our simulation we get the
two following position and velocity update equations using equations
\ref{eq:delta_u} and \ref{eq:delta_p}:
\begin{align*}
    p_v(t+\Delta t) &= p_v(t) + \Delta p_v \\
    u_v(t+\Delta t) &= u_v(t) + \Delta u_v
\end{align*}
We use these equations to perform step 4 of the simulation.

Step 5 is a simple update of the time $t = t + \Delta t$ and step 6 simply make
us loop steps 2 to 5 until we reach the stopping time $t_{stop}$ of the
simulation.

We have now fully described the simple case of simulating a mass-spring system
using the explicit Euler method. The next subsections describe the three
different methods/modifications for applying LAOT to the simulation just
described. The three methods all rely on different ideas to implement LAOT. They
all include using some sort of heuristics in order to perform the adaption of
the methods. We developed and tested these heuristics as we developed the
methods (chronologically) and hence most of the heuristics could potentially be
used across all three methods and not only the methods on which they were used.

\subsection{Adapt-on-switch}
\label{sec:method_switch}
The first LAOT method we propose is the time-adaptive ``Adapt-on-switch''
method. We perform simulation with an initial timestep size. In each step
we try to predict whether a timestep of the current size using the current
velocities will result in a switch of the left and right vertices of any
of the springs of the system. If this is the case, we should try to avoid the switch by
decreasing the timestep of the current iteration, re-calculate the velocity
and position changes and resume stepping of the simulation using the new
timestep size. When switches aren't detected anymore we gradually increase the
timestep of the simulation again.

The switch of two endspoints connected by a spring $s$ can easily be
calculated by checking the length ($p_{l,s} - p_{r,s}$) of the spring before
and after a position update. If the sign of the length differs, the two
points must have switched sides and hence adaption of the timestep should be
performed. The adaption of the timestep can either be done by division of the
current timestep by a constant or by having a finite set of timestep sizes to
choose from.

@@

The heuristics can be chosen to fit the domain of the simulation, but we
have chosen to use the number of steps since last update, $n_{last}$, and
the ratio between the current spring length and the spring length at the
last update $r_{last}$. This should force a bound on the maximal number of
timesteps before update as well as force frequent updates when the spring
is affected by a high acceleration in opposite direction of the current
velocity/direction of change. Looking at the mechanics of a spring we know
that a large force/acceleration is the result of a stretched/contracted
spring. At the ``endpoints'' of the oscillating spring movement the spring
length change decreases slowly followed by a slow increase in the opposite
direction until movement has been accelerated up again. The critical
``endpoints'' of the spring movement should therefore be characterized by a
low change in spring length. Using these observations we should be able to
force updates when $r_{last}$ is low. Further details are explained in section
\ref{sec:experiments_atomic}.

\subsection{Global Estimated Acceleration}
The third and last LAOT method we propose is time-adaptive. The simulations
are done using an initial timestep size and define the timestep of each
iteration based on a heuristic using the acceleration of the springs, and
hence we call the method ``Global Estimated Acceleration'' (GEA). We know
that the force and hence acceleration of a spring is somewhat linear close to
the spring's rest length. Once the spring is streched or contracted away from
the rest length the force ``bends'' off.
Based on this knowledge we will try to make a heuristic telling us about the
needed timestep size at any given point during the simulation. The smallest
needed timestep of the system defines the global timestep for that given step.

For this simple 1D example we use the inverse of the acceleration
for adapting the timestep and hence the heuristic is defined as:
\begin{align}
    \frac{1}{a} &= \frac{1}{\frac{-kx}{m}} = \frac{m}{-kx} \\
    \label{eq:inverse}
    \Delta t &= \text{min}\left( \frac{1}{\abs{a} + 0.1} \right )
\end{align}
We have added a small value (0.1) in order to avoid the timestep to increase
exponentially towards infinity as $\abs{a}$ approaches 0. Furthermore a scaling
of the timestep could be needed depending on the scale of the simulation mesh.
Further details are explained in section \ref{sec:experiments}.

Our though is that for more complex simulations we are able to make estimates
of this type of heuristic less costly than performing the actual simulations
of a small fixed size timestep.

@@

\section{Experiments, results, and discussion hereof}
\label{sec:experiments}

We first give an example of the problem of simulations using the standard
mass-spring simulator leading to the need for application of LAOT. Figure
\ref{fig:std} shows two simulations of the same set of 23 particles distributed
with two different scales simulated for 250 seconds each. All the springs
$s \in S$ are initiated in a stretched state/position with a length 1.25
times longer than the rest length, damping constants $C_s = 0.5$, stiffness
constants $K_s = 0.5$. All vertices $v \in V$ have a mass of $m_v = 30$. Given
this initial state, all springs should start constracting since they are in a
stretched position and eventually somewhere further in time stabilize at their
rest positions since we have added a damping force to the system.

Figure \ref{fig:std_multi_30fps} shows a simulation of the system described above
using 30 fps and figure \ref{fig:std_multi_1fps} shows the simulation using 1
fps. We clearly see that using a timestep of 30 fps the simulation is stable
and able to handle the compression of all the small springs around $t = 110$,
whereas a switch occurs when simulating at 1 fps. This switch causes ripples
of sideeffects for the rest of the simulation and hence the final result is
incorrect. No switches of the longer springs are however occuring and hence
these pose no problem at 1 fps. We will use this simulation for testing all
three LAOT methods proposed in section \ref{sec:methods} and denote it the
multi-scale simulation.

@@

\subsection{Adapt-on-switch}
\label{sec:experiment_switch}
As mentioned in section \ref{sec:method_switch} the Adapt-on-switch method
can be implemented using either a relative or a absolute adaption of the
timestep. In our experiments we have tested out both methods defining the
relative adaption to scale the timestep by a factor of $\frac{1}{100}$ and
the absolute solution to set the timestep to $\frac{1}{30}$ when switches are
detected since this generated nice simulations for the standard simulation.
Furthermore the up-scaling for the relative scheme is set to a factor of 10
since we wish to avoid running into new switches immediately.

Figure \ref{fig:switch} shows the result of running the multi-scale
simulation at 1 fps using the Adapt-on-switch scheme, where the
red crosses mark timesteps at which a switch is predicted in the
following step and hence the schemes are ``activated''. Figure
\ref{fig:switch_multi_1fps_relative} is simulated using the relative scheme
and figure \ref{fig:switch_multi_1fps_absolute} is simulated using the
absolute scheme. As we clearly see in both figures the schemes fail to correct
the simulations in spite of the switch prediction. Once the first switch of
each of the simulations occurs the simulations will change behavior and hence
the first red cross marks the point at which the simulation can be rejected
from being correct.

Taking these observations into account we can safely say that the
Adapt-on-switch method/scheme is not working for our multi-scale simulation. This does
however not mean that the method couldn't work for other domains or other
refinements of the method. The problem with the scheme is that the prediction
of switches is done too late in the simulation. Once we detect the switches
it is too late to do anything about them given our model of the mass-spring
system. In order for the scheme to work properly it should be modified to be
able to perform the detection of switches earlier in the simulation if such a
prediction can be made reliably.

In the real world the spring force would ``bend'' towards $\pm \infty$ once
the spring is stretched or contracted too much resulting in a maximum and
minimum spring length. In other words the linear model that we use does not
suffice when using the Adapt-on-switch scheme, since the spring forces aren't
reacting the way we need them to when a spring is contracted. Having used a
more realistic choice of model could possibly have had a positive impact of
the use of this scheme. This would cause the spring force of the spring
in question to grow in order to avoid the switch.

As a sidenote it should be mentioned that we very shortly tested out an
alternative way of calculating the adaptive stepsize to confirm the hypothesis
above. We predicted how long a timestep we could take before the switch would
occur and then divided this steplength by 2 to get a new timestep length. This
generated infeasibly small timesteps forcing us to stop the simulation before
the time limit $t_{stop}$ was reached. This supports our hypothesis of the
incompatibility between the domain and the Adapt-on-switch method in the current
state.

@@

\subsection{Atomic Updating using Heuristics}
\label{sec:experiments_atomic}
The AUH method concists of two heuristics for determining whether an element
should be updated or not: steps since last update $n_{last}$, and the ratio
$r_{last}$ between the spring length in last update and the current length.

We first test the method using only the first heuristic which will give us
synchronized updates of all vertices every $r_{last}$. step. Using the same
setup as the multi-scale simulation we perform simulations using the AUH method
with the length ratio heuristic ignored, the timestep set to $\Delta t =
\frac{1}{10}$ (10 fps) and varing $n_{last}$ across three different simulations.
Figure \ref{fig:atomic_multi_sbound} (\ref{fig:atomic_multi_10sbound} to
\ref{fig:atomic_multi_69sbound}) shows the resulting simulations where
figures \ref{fig:atomic_multi_10sbound}, \ref{fig:atomic_multi_68sbound},
and \ref{fig:atomic_multi_69sbound} uses a $n_{last}$ of 10, 68, and 69
respectively.

Simulating using 10 fps and $n_{last} = 10$ give us the same amount of
velocity updates as the standard simulation using 1 fps (which had switches).
Using the AUH scheme we do however get a nice and smooth simulation with no
problems. As can be seen in the figures \ref{fig:atomic_multi_68sbound} and
\ref{fig:atomic_multi_69sbound} we are able to perform the simulation using
$n_{last} = 68$ without having switches occuring whereas using $n_{last} =
69$ results in switches occuring. In other words we are able to perform the
simulation at 10 fps but with 6.8 seconds between acceleration and velocity
calculations. This is a result of combining the explicit Euler method with
the AUH scheme. The scheme extrapolates the position at the last
acceleration and velocity update to a position close in time to the next
update position. It then uses this position when updating the acceleration
and velocity again. Had we chosen an alternative time integration method this
advantage would most likely have been eliminated and let us perform the larger
timesteps with the standard simulation. This would however had increased the
computational cost of each timestep.

The second heuristic ($r_{last}$) should be activated when $r_{last}$ is
small. The bound $r_{b}$ for forcing updates based on $r_{last}$ should
however be defined according to the simulation in question. Likewise the
first heuristic should be activated in between the activation of the $r_b$
bound to force a minimum of updates between the endpoints of the spring
motion.

In order to test out the heuristic we first perform a simulation using
only 3 equally distributed particles at 10 fps for 250 seconds. The initial
state of the simulation is: timestep $\Delta t = 0.1$, damping $C_s = 0.01$,
stiffness $K_s = 0.5$, initial vertex positions 2, 0 and -2, and spring rest
length 1.6 for both springs. By performing a series of simulations we find
the most optimal combination of the two heuristics for the simulation to be
$r_{last} = 20$, $r_b = 1.002$. Figure \ref{fig:atomic_uniform} shows the
resulting simulation plots displaying positions as a function of the time
using the AUH method (\ref{fig:atomic_uniform_10fps_0}) and the standard
method (\ref{fig:atomic_uniform_10fps_1}). The green dots of the AUH plot
shows the update positions.

As we see from the AUH plot, the heuristics start out by updating the
accelerations and velocities at the endpoints of the spring motions as
intended. Likewise the stepping bound activates updates at lower frequency
(every 20th step) in between the endpoints as intended. When time passes the
system is being dampened too much compared to the small damping factor. At the
end of the simulation the $r_b$ bound activates updates in all steps of the
simulation. When we compare the resulting simulation of the AUH and standard
methods we are observing quite large differences. As mentioned the AUH method
dampens the system too much when the $r_b$ bound is not activated, and hence we
get the observed difference in simulation result.

Figure \ref{fig:atomic_multi_both} shows the resulting simulation plot
using the AUH method and setting $n_{last} = 10,~r_b = 1.01$ for the
multi-scale simulation. As for the previous simulation example we have
estimated the best values of $n_{last}$ and $r_b$ by varying both and
inspecting the resulting simulation. A switch occurs at time $t = 104$ which
corrupts and invalidates the rest of the simulation.

The conclusion of the application of this method is, that the first heuristic
was very succesful in case we need to use the explicit Euler method. The second
heuristic was however only affecting the outcome of the simulations negatively,
and thus it should be rejected.
